NMRlipids IV: Headgroup \& glycerol backbone structures, and cation binding in bilayers with PS lipids

### 2019-02-05 Manuscript/manuscriptPS.tex
Style edits in experimental methods; asked for a few clarifications.

\subsection{C-H bond order parameters from the natural abundance $^{13}$C NMR}

Headgroup and glycerol backbone C-H bond order parameters of POPS
were determined from the chemical-shift resolved dipolar splittings
measured with a R-type Proton Detected Local Field (R-PDFL) experiment~\cite{dvinskikh04}.
The corresponding order parameter signs were measured with a S-DROSS experiment~\cite{gross97}
using natural abundance $^{13}$C solid state NMR spectroscopy as described previously \cite{ferreira13,ferreira16}.
The experiments were done in a Bruker Avance III 400 spectrometer operating at a $^1$H Larmor frequency of 400.03 MHz.
Magic angle spinning (MAS) of the sample was used at a frequency of 5.15 kHz (R-PDLF experiment) and 5 kHz (S-DROSS experiment).
The following experimental setups were used.

{\emph{C-H bond order parameters from the R-PDLF experiment}}. The parameters are described according to Figures 1c and 2c of the original reference
for the R-PDLF experiment~\cite{dvinskikh04}.  The refocused-INEPT delays $\tau_1$ and $\tau_2$ were 1.94 ms and 0.97 ms, respectively.
Radio frequency pulses with the nutation frequencies: 46.35 kHz (R18$^7_1$ pulses), 63.45 kHz ($^{13}$C 90$^{\rm{o}}$ and 180$^{\rm{o}}$),
50 kHz (SPINAL64 $^1$H decoupling pulses). The $t_1$ increment was equal to 10.79 $\mu$s $\times18\times2$ and 32 points in the indirect
dimension were recorded using 1024 scans for each, with recycle delay of 5 s and a spectral with of 149.5 ppm.

\emph{Order parameter signs from the S-DROSS experiment}.
The parameters are described according to Figures 1b and 1c of the original reference for the S-DROSS
experiment~\cite{gross97}. The refocused-INEPT delay $\delta_2$ was 1.19 ms. The $\tau_1$ and $\tau_2$ in the S-DROSS recoupling
blocks $R$ were set as 39.4 $\mu$s and 89.4 $\mu$s, respectively. Radio frequency pulses with the nutation
frequencies: 63.45 kHz ($^{13}$C 90$^{\rm{o}}$ and 180$^{\rm{o}}$), 50 kHz ($^1$H SPINAL64 decoupling).
The $t_1$ increment (dipolar recoupling dimension) was 800 $\mu$s and a total of 8 points along $t_1$ were
measured using 1024 scans for each with a recycle delay of 5 s and a spectral with of 149.5 ppm.

\emph{Numerical simulations of S-DROSS curves.}
The numerical simulations of S-DROSS curves were performed with the SIMPSON simulation package \cite{bak00} using the $^{13}$C--$^1$H
dipolar couplings determined by the R-PDLF experiments or calculated from the known $^2$H quadrupolar couplings \cite{browning80} as input.
The chemical shift anisotropy and homonuclear couplings were neglected, and the input file {\it{rep2000}} was used to simulate the random
distribution of bilayer orientations in the samples studied.

\emph{Sample preparation}
The sample was prepared simply by mixing the POPS with water (lipid:water 60:40 wt\%) in an eppendorf
tube mixing and centrifuging the sample repeatedly until an homogeneous viscous fluid was obtained.
20 mg of sample was then transferred to an NMR insert suitable for 4 mm NMR rotors.  
\todo{Maybe we need little bit more information about the mixing procedure?}

@@

\begin{figure}[!tb]
  \centering
  \includegraphics[width=8.5cm]{../Figs/fig1_POPS.pdf}
  \caption{\label{PShgSIGNSsimpson}
    The headgroup and glycerol backbone region of the (A) INEPT spectrum and
    (B) 2D R-PDLF spectra.
    (C) Experimental S-DROSS data (points), and SIMPSON simulations (blue lines) with
    the order parameter values of -0.12 for the $\beta$-carbon, and 0.09 and -0.02
    for the $\alpha$-carbon splittings.
    Dashed gray line is the S-DROSS curve from a SIMPSON simulation with a positive value (+0.02) 
    for the smaller $\alpha$-carbon order parameter.
  }
  \todo{I think that the peak labeling would be good to show also in (A).}\\
  \todo{Also for $\alpha$ these are OPs, right,  thus we could just exclude the word 'splittings'?}
  \todo{Please confirm: Was the positive value used for $\alpha$ in the gray curve 0.02?}
\end{figure}
\begin{figure}[!htb]
  \centering
  \includegraphics[width=9.0cm]{../Figs/PCPScomp.eps}
  \caption{\label{HGorderParameters}
    (A) Chemical structures and labels for the headgroup and glycerol backbone carbons.
    (B) Headgroup and glycerol backbone order parameters of POPS measured in this work compared
    with the previously published values from DOPS ($^2$H NMR, 0.1M of NaCl) \cite{browning80} and 
    POPC  ($^{13}$C NMR) \cite{ferreira13} experiments. Signs of the PS order parameters
    are measured in this work. Signs of the PC order parameters are measured previously~\cite{ferreira16}.
  }\todo{
  1) Use diamonds for DOPS, spheres for POPS.
  2) Error bars?
  3) Change the y-label to $S_\mathrm{CH}$, and invert the y-axis, as in Fig. 3.
  4) Make lower g1 POPC visible, e.g., by slightly larger point.
  }
\end{figure}

The INEPT and 2D R-PDLF experiments from POPS sample give well resolved spectra for all the
carbons in the headgroup and glycerol backbone regions (Fig. \ref{PShgSIGNSsimpson}).
The glycerol backbone carbon peaks were assigned according to the POPC spectra~\cite{ferreira13}.
The peaks for $\beta$ and $\alpha$ carbons were assigned according to the
known order parameters from the $^2$H\,NMR experiments~\cite{browning80}.
Slices of the R-PDLF spectra and the resulting order parameter values
are shown in the supplementary information (Fig. \ref{DPslices}). 
Since the R-PDLF and previous $^2$H\,NMR experiments \cite{browning80,roux91} give 
only the absolute values of order parameters, we determined the signs of the PS headgroup
order parameters using the S-DROSS experiment \cite{gross97}.
The S-DROSS slice clearly shows that the order parameter of
the $\beta$-carbon is negative (Fig. \ref{PShgSIGNSsimpson} C)), \todo{
Could we explain in a few words how this is clearly seen?}
which is confirmed by SIMPSON simulations. The beginning of the S-DROSS slice
suggests that the larger order parameter of the $\alpha$-carbon 
is positive and the deviation towards negative values with longer T$_1$ times suggests
that the smaller order parameter is negative. This is confirmed by a SIMPSON simulation
using the value of -0.02 from $^2$H\,NMR experiment \cite{roux91} for the smaller order parameter.
The literature value was used because the
resolution of our experiment was not sufficient to determine the
small value of the order parameter.
The S-DROSS curve from SIMPSON simulation with a positive value for the smaller order parameter
(dashed grey in Fig. \ref{PShgSIGNSsimpson} C)) did not agree with the experiment, 
confirming \todo{Or rather 'corroborating'?}
the interpretation that the smaller order parameter is negative.

@@

The headgroup and glycerol backbone order parameters of 
POPS measured in this work are in good agreement with the previously reported
values from $^2$H\,NMR experiments of DOPS \cite{browning80} (Fig. \ref{HGorderParameters}).
When compared with the previously measured values for POPC \cite{ferreira13} (Fig. \ref{HGorderParameters}),
the $\beta$-carbon order parameter is significantly more negative and $\alpha$-carbon
experiences a significant forking (different order parameters for the two hydrogens in the same carbon \cite{ollila16}) in the PS headgroup.
These features have been intepreted to arise from a rigid PS headgroup
conformation, stabilized by hydrogen bonds or electrostatic
interactions \cite{browning80,buldt81}, but detailed structrural interpretation is not
available.

>>>>

\subsection{C--H bond order parameters from the natural abundance $^{13}$C\,NMR}

Headgroup and glycerol backbone C--H bond order parameters of POPS
were determined from the chemical-shift resolved dipolar splittings
measured with a R-type Proton Detected Local Field (R-PDLF) experiment~\cite{dvinskikh04}.
The corresponding order parameter signs were measured with a S-DROSS experiment~\cite{gross97}
using natural abundance $^{13}$C solid state NMR spectroscopy as described previously \cite{ferreira13,ferreira16}.
The experiments were done in a Bruker Avance III 400 spectrometer operating at a $^1$H Larmor frequency of 400.03 MHz.
Magic angle spinning (MAS) of the sample was used at a frequency of 5.15 kHz (R-PDLF experiment) and 5 kHz (S-DROSS experiment).
The following experimental setups were used.

{\emph{C--H bond order parameters from the R-PDLF experiment.}} The parameters are described according to Figures 1c and 2c of the original reference
for the R-PDLF experiment~\cite{dvinskikh04}.  The refocused-INEPT delays were $\tau_1=$\,1.94\,ms and $\tau_2=$\,0.97\,ms.
Radio frequency pulses with the nutation frequencies: 46.35 kHz (R18$^7_1$ pulses), 63.45 kHz ($^{13}$C 90$^{\rm{o}}$ and 180$^{\rm{o}}$),
50 kHz (SPINAL64 $^1$H decoupling pulses).
\todo{Verb missing. Would it be correct to say: 'Radio frequency pulses had the nutation frequencies:'?}
The $t_1$ increment was equal to 10.79 $\mu$s $\times18\times2$, and 32 points in the indirect
dimension were recorded using 1024 scans for each, with a recycle delay of 5\,s and a spectral with \todo{'with' $\to$ 'width'?} of 149.5 ppm.

\emph{Order parameter signs from the S-DROSS experiment.}
The parameters are described according to Figures 1b and 1c of the original reference for the S-DROSS
experiment~\cite{gross97}. The refocused-INEPT delay $\delta_2$ was 1.19 ms. The $\tau_1$ and $\tau_2$ in the S-DROSS recoupling
blocks $R$ were set as $\tau_1=$\,39.4\,$\mu$s and $\tau_2=$\,89.4 $\mu$s. Radio frequency pulses with the nutation
frequencies: 63.45 kHz ($^{13}$C 90$^{\rm{o}}$ and 180$^{\rm{o}}$), 50 kHz ($^1$H\,SPINAL64 decoupling).
\todo{Verb missing. Would it be correct to say: 'Radio frequency pulses had the nutation frequencies:'?}
The $t_1$ increment (dipolar recoupling dimension) was 800 $\mu$s, and a total of 8 points along $t_1$ were
measured using 1024 scans for each, with a recycle delay of 5\,s and a spectral with \todo{'with' $\to$ 'width'?}  of 149.5 ppm.

\emph{Numerical simulations of S-DROSS curves.}
The numerical simulations of S-DROSS curves were performed with the SIMPSON simulation package \cite{bak00}
using as input the $^{13}$C--$^1$H
dipolar couplings, either as determined by the R-PDLF experiments, or as calculated from the known $^2$H quadrupolar couplings \cite{browning80}.
The chemical shift anisotropy and homonuclear couplings were neglected, and the input file {\it{rep2000}} was used to simulate a random
distribution of bilayer orientations in the samples studied.

\emph{Sample preparation.}
The sample was prepared simply by mixing the POPS \todo{From some company?} with water (lipid:water 60:40\,wt-\%) in an Eppendorf
tube by mixing and centrifuging the sample repeatedly until a \todo{visually(?)} homogeneous viscous fluid was obtained.
Then 20 mg of the sample was  transferred to an NMR insert suitable for 4 mm NMR rotors.  
\todo{Maybe we need little bit more information about the mixing procedure?}

@@

\begin{figure}[!tb]
  \centering
  \includegraphics[width=8.5cm]{../Figs/fig1_POPS.pdf}
  \caption{\label{PShgSIGNSsimpson}
    The headgroup and glycerol backbone region of the (A) INEPT spectrum and
    (B) 2D R-PDLF spectra.
    (C) Experimental SDROSS data (points) and SIMPSON simulations (blue lines) with
    the order parameter values of -0.12 for the $\beta$-carbon, and 0.09 and -0.02
    for the $\alpha$-carbon slittings. The S-DROSS curve from SIMPSON simulation with positive value
    for the smaller $\alpha$-carbon order parameter (dashed grey).
  }
  \todo{I think that the peak labeling would be good to show also in (A).}
\end{figure}
\begin{figure}[!htb]
  \centering
  \includegraphics[width=9.0cm]{../Figs/PCPScomp.eps}
  \caption{\label{HGorderParameters}
    (A) Chemical structures and labels for the headgroup and glycerol backbone carbons.
    (B) Headgroup and glycerol backbone order parameters of POPS measured in this work compared
    with the previously published values from DOPS ($^2$H NMR, 0.1M of NaCl) \cite{browning80} and 
    POPC  ($^{13}$C NMR) \cite{ferreira13} experiments. Signs of the PS order parameters
    are measured in this work. Signs of the PC order parameters are measured previously~\cite{ferreira16}.
  }
\end{figure}

The INEPT and 2D R-PDLF experiments from POPS sample give well resolved spectras for all the
carbons in headgroup and glycerol backbone region (Fig. \ref{PShgSIGNSsimpson}).
The glycerol backbone carbon peaks were assigned according to the POPC spectra~\cite{ferreira13}.
The peaks for $\beta$ and $\alpha$ carbons were assigned according to the
known order parameters from the $^2$H NMR experiments~\cite{browning80}.
Slices of the R-PDFL spectra and the resulting order parameters values
are shown in the supplementary information (Fig. \ref{DPslices}). 
Since the R-PDFL and previous $^2$H NMR experiments \cite{browning80,roux91} give 
only the absolute values of order parameters, we determined the signs of PS headgroup
order parameters using the S-DROSS experiment \cite{gross97}.
The S-DROSS slice clearly shows that the order parameter of
the $\beta$-carbon is negative (Fig. \ref{PShgSIGNSsimpson} C)),
which is confirmed by SIMPSON simulations. The beginning of the S-DROSS slice
suggests that the larger order parameter of the $\alpha$-carbon 
is positive and the deviation towards negative values with longer T$_1$ times suggests
that the smaller order parameter is negative. This is confirmed by a SIMPSON simulation
using the value of -0.02 from $^2$H NMR experiment \cite{roux91} for the smaller order parameter.
The literature value was used because the
resolution of our experiment was not sufficient to determine the
small value of the order parameter.
The S-DROSS curve from SIMPSON simulation with a positive value for the smaller order parameter
(dashed grey in Fig. \ref{PShgSIGNSsimpson} C)) did not agree with the experiment, 
confirming the interpretation that the smaller order parameter is negative.

@@

The headgroup and glycerol backbone order parameters of 
POPS measured in this work are in good agreement with the previously reported
values from $^2$H NMR experiments of DOPS \cite{browning80} (Fig. \ref{HGorderParameters}).
When compared with the previously measured values for POPC \cite{ferreira13} (Fig. \ref{HGorderParameters}),
the $\beta$-carbon order parameter is significantly more negative and $\alpha$-carbon
experiences a significant forking (different order parameters for two hydrogens in the same carbon \cite{ollila16}) in the PS headgroup.
These features have been intepreted to arise from a rigid PS headgroup
conformation, stabilized by hydrogen bonds or electrostatic
interactions \cite{browning80,buldt81}, but detailed structrural interpretation is not
available.
